Bow shocks, bow waves, and dust waves. IV. Bow shape statistics

### 2019-04-07 papers/Papers1234/sec-strong-coupling.tex
Correct typo in eq (24) of Paper I
Update Paper I to split shell profiles figure in 3 parts

  x^3 \tau_* > \frac{4 S c \sound \bar{m}^2 \kappa}{3 \alpha L} \ .

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/shell-profiles}
  \caption{Profiles of physical variables along the axis of bow shells
    for (a)~wind-supported bow shock; (b)~radiation-supported bow
    wave, and (c)~radiation-supported bow shock.  In all cases, a
    stream velocity of \SI{30}{km.s^{-1}} and an equilibrium ionized
    sound speed of \SI{10}{km.s^{-1}} are assumed.}
  \label{fig:axial-structure}
\end{figure}

In the shell of a wind-supported bow shock, the gas density is highest
at the contact discontinuity, but this is no longer true for
radiation-supported bows, as illustrated in
Figure~\ref{fig:axial-structure}, where we schematically show the
shell structure for the three cases: WBS, RBW, and RBS.  In all cases,
we assume a far-field mach number of \(\M_\infty = 3\) for the incident
stream of density \(n_\infty\).  For the WBS case
(Fig.~\ref{fig:axial-structure}a), the stream is unchanged until it
reaches the shock, so \(\M_0 = \M_\infty\) and \(n_0 = n_\infty\).  The density
increases to \(n_1 = 3 n_0\) in the shock
(eq.~[\ref{eq:shock-n-jump}]), and then to \(n_2 = 9 n_0\)
(eq.~[\ref{eq:isothermal-shell-density}]) after it has cooled back
down to the equilibrium photoionized temperature.  In the figure, we
show the case where \(d\cool < h_0\), so that most of the shell is at
roughly constant density.  Note that although the mass flux along the
axis, \(n v\), is approximately conserved in most of the flow, this is
no longer true close to the contact discontinuity, since the
streamlines bend away from the axis due to lateral pressure gradients.
This is what allows \(v\) to tend to zero, while \(n\) remains
constant.  The thermal pressure in the shocked stellar wind is equal
to the shell pressure, but its density is much smaller due to a
temperature that is higher by a factor of \(\approx 3100 V_3^2\),
(eq.~[\ref{eq:shock-T-jump}], assuming inefficient cooling).

For the optically thin RBW case (Fig.~\ref{fig:axial-structure}b),
equations~\eqref{eq:rad-accel} to~\eqref{eq:rad:R0} can be combined to
give the radial dependence of the stream velocity as it approaches the
star:
\begin{equation}
  \label{eq:v-RBW-ballistic}
  v(R) = \left( 1 - \frac{R_{**}}{R} \right)^{1/2} v_\infty \ .
\end{equation}
This ballistic approximation (shown by the dashed red line in
Fig.~\ref{fig:axial-structure}b) predicts that the velocity smoothly
decreases to zero at \(R = R_{**}\).  However, although this is valid
in the limit of high Mach number, it ignores the effects of gas
pressure and so becomes invalid once the stream velocity falls close
to the sound speed.  The situation bears similarities to the flow in a
supersonic diffuser, such as the inlet of a ramjet or other supersonic
aircraft engine \citep{Seddon:1999a}, which slows supersonic airflow
down to subsonic speeds before combustion.  Exactly the opposite flow
configuration is present in a rocket nozzle \citep{Courant:1948a},
where the flow can smoothly pass from subsonic to supersonic
velocities at the throat of the nozzle.  An analogy between an
isothermal stellar wind and a rocket nozzle is developed in detail in
\S~3.5 of \citet{Lamers:1999b}.  However, in the reverse case of a
supersonic diffuser, a smooth transition from supersonic to subsonic
flow is not possible \citep{Morawetz:1956a}.  Instead, a normal shock
wave develops, which decelerates the supersonically entering flow,
allowing it to exit subsonically \citep{Embid:1984a, Hafez:1999a}.
The supersonic inlet/bow wave analogy is not so exact as the rocket
nozzle/stellar wind analogy due to the multi-dimensional nature of the
bow shell, which means that lateral flows of gas away from the apex
become important when the velocity is subsonic.\footnote{%
  In the aeronautical case, the inlet is usually designed so that
  multiple oblique shocks partially decelerate the flow before it
  passes through the normal shock.  This is the analogy of the initial
  gradual radiative deceleration in the bow wave.  } %
Nonetheless, we expect that a normal shock must be present in the
flow, although there does not seem to be any simple argument for
predicting the exact Mach number where it will occur.  In
Figure.~\ref{fig:axial-structure}b we assume that the shock occurs at
a Mach number \(\M_0 = 1.5\), but multidimensional numerical
simulations are necessary to test this supposition.

@@

For the optically thick RBS case (Fig.~\ref{fig:axial-structure}c),
the shell density profile is no longer simply exponential since the
pressure scale height now increases as one moves away from the star,
due to the extinction-induced decline in radiation flux. Assuming
single scattering and plane geometry, the resultant hydrostatic
density profile is doubly exponential:
\begin{equation}
  \label{eq:rbs-density-profile}
  n(R) = n_2 e^{-e^{-x}} \text{\quad where\quad } x = n_2 \bar{m} \kappa (R - R_*) \ . 
\end{equation}
The shell density increases from \(e^{-1} n_2\) at \(R_*\) to saturate
at \(n_2\) for \(R > R_* + h_a\), where \(h_a\) is now the pressure
scale height at the inner edge of the shell.  From
equations~(\ref{eq:Rstar}, \ref{eq:rad:R0}, \ref{eq:tau-thin},
\ref{eq:strong-cooling-h0}, \ref{eq:scale-height}), and taking \(R_0 \approx R_*\), we find
\begin{equation}
  \label{eq:ha-versus-h0}
  h_a \approx \frac{4 h_0}{3 \tau} \ .
\end{equation}
We show results for a total shell optical depth \(\tau = 3\), for which
\(h_a\) is a significant fraction of \(h_0\).  For much larger optical
depths, \(h_0 \gg h_a\) so that the constant density portion of the
shell will dominate the total column.  A small fraction \(e^{-\tau}\) of
the photons will penetrate the shell and be available to decelerate
the supersonic stream, as in the RBW case, yielding
\(v_0 = (1 - \tau e^{-\tau})\, v_\infty\).  In the illustrated case of
\(\tau = 3\), the resultant speed reduction before the shock is only 8\%.
As mentioned in \S~\ref{sec:effects-gravity}, radiative acceleration
becomes less important than gravity in the outer regions of an opaque
shell.  However, in order for this to have a significant effect on the
shell, the gravitational scale height would have to be be less than
the shell thickness: \(\Gamma\Edd h_a < h_0\).  From
equation~\eqref{eq:ha-versus-h0} this translates to
\(\tau > \frac43 \Gamma\Edd\), where dust-opacity Eddington factors are
\(\Gamma\Edd \sim 100\) for OB stars.  Such large optical depths would
require extremely high ambient densities of
\(n > \num{2e7} L_4 \,v_{10}^2 \left( \SI{10}{M_\odot}/M \right)^2 \,
\si{cm^{-3}} \), where we have used equations~(\ref{eq:tau-star},
\ref{eq:eddington-factor}) and that \(\tau \approx 2\tau_*\) in the RBS limit.

>>>>

  x^3 \tau_* > \frac{4 S c \sound^2 \bar{m}^2 \kappa}{3 \alpha L} \ .

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/shell-profile-wbs}
  \caption{Profiles of physical variables along the axis of the bow
    shell for a wind-supported bow shock.  The case is shown of a
    stream velocity of \SI{30}{km.s^{-1}} and an equilibrium ionized
    sound speed of \SI{10}{km.s^{-1}}.}
  \label{fig:axial-structure-wbs}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/shell-profile-rbw}
  \caption{As Fig.~\ref{fig:axial-structure-wbs} but for a
    radiation-supported bow wave.}
  \label{fig:axial-structure-rbw}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/shell-profile-rbs}
  \caption{As Fig.~\ref{fig:axial-structure-wbs} but for a
    radiation-supported bow shock.}
  \label{fig:axial-structure-rbs}
\end{figure}

In the shell of a wind-supported bow shock, the gas density is highest
at the contact discontinuity, but this is no longer true for
radiation-supported bows, as illustrated in
Figures~\ref{fig:axial-structure-wbs} to
\ref{fig:axial-structure-rbs}, where we schematically show the shell
structure for the three cases: WBS, RBW, and RBS.  In all cases, we
assume a far-field mach number of \(\M_\infty = 3\) for the incident stream
of density \(n_\infty\).  For the WBS case
(Fig.~\ref{fig:axial-structure-wbs}), the stream is unchanged until it
reaches the shock, so \(\M_0 = \M_\infty\) and \(n_0 = n_\infty\).  The density
increases to \(n_1 = 3 n_0\) in the shock
(eq.~[\ref{eq:shock-n-jump}]), and then to \(n_2 = 9 n_0\)
(eq.~[\ref{eq:isothermal-shell-density}]) after it has cooled back
down to the equilibrium photoionized temperature.  In the figure, we
show the case where \(d\cool < h_0\), so that most of the shell is at
roughly constant density.  Note that although the mass flux along the
axis, \(n v\), is approximately conserved in most of the flow, this is
no longer true close to the contact discontinuity, since the
streamlines bend away from the axis due to lateral pressure gradients.
This is what allows \(v\) to tend to zero, while \(n\) remains
constant.  The thermal pressure in the shocked stellar wind is equal
to the shell pressure, but its density is much smaller due to a
temperature that is higher by a factor of \(\approx 3100 V_3^2\),
(eq.~[\ref{eq:shock-T-jump}], assuming inefficient cooling).

For the optically thin RBW case (Fig.~\ref{fig:axial-structure-rbw}),
equations~\eqref{eq:rad-accel} to~\eqref{eq:rad:R0} can be combined to
give the radial dependence of the stream velocity as it approaches the
star:
\begin{equation}
  \label{eq:v-RBW-ballistic}
  v(R) = \left( 1 - \frac{R_{**}}{R} \right)^{1/2} v_\infty \ .
\end{equation}
This ballistic approximation (shown by the dashed red line in
Fig.~\ref{fig:axial-structure-rbw}) predicts that the velocity smoothly
decreases to zero at \(R = R_{**}\).  However, although this is valid
in the limit of high Mach number, it ignores the effects of gas
pressure and so becomes invalid once the stream velocity falls close
to the sound speed.  The situation bears similarities to the flow in a
supersonic diffuser, such as the inlet of a ramjet or other supersonic
aircraft engine \citep{Seddon:1999a}, which slows supersonic airflow
down to subsonic speeds before combustion.  Exactly the opposite flow
configuration is present in a rocket nozzle \citep{Courant:1948a},
where the flow can smoothly pass from subsonic to supersonic
velocities at the throat of the nozzle.  An analogy between an
isothermal stellar wind and a rocket nozzle is developed in detail in
\S~3.5 of \citet{Lamers:1999b}.  However, in the reverse case of a
supersonic diffuser, a smooth transition from supersonic to subsonic
flow is not possible \citep{Morawetz:1956a}.  Instead, a normal shock
wave develops, which decelerates the supersonically entering flow,
allowing it to exit subsonically \citep{Embid:1984a, Hafez:1999a}.
The supersonic inlet/bow wave analogy is not so exact as the rocket
nozzle/stellar wind analogy due to the multi-dimensional nature of the
bow shell, which means that lateral flows of gas away from the apex
become important when the velocity is subsonic.\footnote{%
  In the aeronautical case, the inlet is usually designed so that
  multiple oblique shocks partially decelerate the flow before it
  passes through the normal shock.  This is the analogy of the initial
  gradual radiative deceleration in the bow wave.  } %
Nonetheless, we expect that a normal shock must be present in the
flow, although there does not seem to be any simple argument for
predicting the exact Mach number where it will occur.  In
Figure.~\ref{fig:axial-structure-rbw} we assume that the shock occurs at
a Mach number \(\M_0 = 1.5\), but multidimensional numerical
simulations are necessary to test this supposition.

@@

For the optically thick RBS case (Fig.~\ref{fig:axial-structure-rbs}),
the shell density profile is no longer simply exponential since the
pressure scale height now increases as one moves away from the star,
due to the extinction-induced decline in radiation flux. Assuming
single scattering and plane geometry, the resultant hydrostatic
density profile is doubly exponential:
\begin{equation}
  \label{eq:rbs-density-profile}
  n(R) = n_2 e^{-e^{-x}} \text{\quad where\quad } x = n_2 \bar{m} \kappa (R - R_*) \ . 
\end{equation}
The shell density increases from \(e^{-1} n_2\) at \(R_*\) to saturate
at \(n_2\) for \(R > R_* + h_a\), where \(h_a\) is now the pressure
scale height at the inner edge of the shell.  From
equations~(\ref{eq:Rstar}, \ref{eq:rad:R0}, \ref{eq:tau-thin},
\ref{eq:strong-cooling-h0}, \ref{eq:scale-height}), and taking \(R_0 \approx R_*\), we find
\begin{equation}
  \label{eq:ha-versus-h0}
  h_a \approx \frac{4 h_0}{3 \tau} \ .
\end{equation}
We show results for a total shell optical depth \(\tau = 3\), for which
\(h_a\) is a significant fraction of \(h_0\).  For much larger optical
depths, \(h_0 \gg h_a\) so that the constant density portion of the
shell will dominate the total column.  A small fraction \(e^{-\tau}\) of
the photons will penetrate the shell and be available to decelerate
the supersonic stream, as in the RBW case, yielding
\(v_0 = (1 - \tau e^{-\tau})\, v_\infty\).  In the illustrated case of
\(\tau = 3\), the resultant speed reduction before the shock is only 8\%.
As mentioned in \S~\ref{sec:effects-gravity}, radiative acceleration
becomes less important than gravity in the outer regions of an opaque
shell.  However, in order for this to have a significant effect on the
shell, the gravitational scale height would have to be be less than
the shell thickness: \(\Gamma\Edd h_a < h_0\).  From
equation~\eqref{eq:ha-versus-h0} this translates to
\(\tau > \frac43 \Gamma\Edd\), where dust-opacity Eddington factors are
\(\Gamma\Edd \sim 100\) for OB stars.  Such large optical depths would
require extremely high ambient densities of
\(n > \num{2e7} L_4 \,v_{10}^2 \left( \SI{10}{M_\odot}/M \right)^2 \,
\si{cm^{-3}} \), where we have used equations~(\ref{eq:tau-star},
\ref{eq:eddington-factor}) and that \(\tau \approx 2\tau_*\) in the RBS limit.
